GPU-Accelerated Parallelization of Edge-Based Motion Detection Algorithms with the NVIDIA CUDA Framework

### 2015-12-02 documents/paper/paper.tex
Add details on CUDA implementation of separable convolution

\documentclass[journal]{IEEEtran}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{algorithm2e}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{float}
\usepackage{bm}

@@

\subsection{Separable Convolution}
The filtering necessary in steps (1) and (2) from our algorithm above involve the convolution of a two-dimensional image and a two-dimensional kernel filter of constant width and height. A na\"{\i}ve implementation of two-dimensional convolution would require a double sum across the kernel and image for each individual pixel:
\begin{equation}
	\label{2d-naive-convolution}
	y[m, n] = x[m, n] * h[m, n] = \sum_{i = -\infty}^{\infty} \sum_{j = -\infty}^{\infty} h[i, j] x[m - i, n - j]
\end{equation}
For an image of size $M$ x $N$ and kernel of size $k$, a direct implementation would require $O(MNk^2)$ time, which is infeasible for a real-time implementation. Instead, it is possible to exploit the separable property of certain convolution matrices $h$, e.g. that $h$ can be represented as the product of two one-dimensional matrices $h_1$ and $h_2$. This effectively reduces our two-dimensional convolution of \eqref{2d-naive-convolution} to two separate instances of one-dimensional convolution:
$$y[m, n] = h_1[n] * (h_2[m] * x[m, n])$$
which is implemented as
$$y[m, n] = \sum_{i = -\infty}^{\infty}h_1[i] \sum_{j = -\infty}^{\infty}h_2[j] x[m - i, n - j]$$
For the same $M$ x $N$ input and square kernel of size $k$, a separable implementation reduces the computational complexity to $O(MNk)$. By comparison, an implementation using the FFT would cost $O(MN\log{MN})$. In practice, consider the following results\footnote{Results were obtained by measuring the execution time of a MATLAB script on an Intel Core i7 4790K @ 4.7 GHz.} for a convolution of a $3$ x $3$ kernel with an image of size $2500$ x $2500$:
\begin{table}[H]
	\small
	\centering
	\caption{Comparison of Computational Complexity of Different Convolution Implementation Algorithms}
	\label{convolution-complexity-comparison}
	\begin{tabular}{llll}
	 & Complexity & Execution time (s) \\
	 Direct convolution & $O(MNk^2)$ & $3$ \\
	 Separated convolution & $O(MNk)$ & $3$ \\
	 FFT & $O(MN\log{MN})$ & $3$
	\end{tabular}
\end{table}
As detailed in later sections, we take advantage of the fast computational complexity of separable convolution in our implementation, as all of the filters we apply are separable into one-dimensional matrices.

@@

\subsection{GPU Parallelization and CUDA Implementation}
We begin with a parallelized implementation of separable convolution, since the same operation is applied for both the initial Gaussian low pass filter

@@

\subsection{Motion Area Estimation via a Spatial Difference Density Map}
The difference matrix $\boldsymbol{D}$ determined with Algorithm \ref{difference-matrix-algorithm} is effectively a maximally precise estimate of motion area (e.g., to the resolution of a single pixel, since $\boldsymbol{D}$ is computed on a pixel-by-pixel basis). However, in practice, it is more useful to consider general regions of the image around which motion exists. For this reason, we build a density map of the matrix $\boldsymbol{D}$ using equal-sized rectangular sections of the input image that approximates the density of the difference per unit area (units pixels$^2$). Thus, the original $M$ x $N$ difference matrix $\boldsymbol{D}$ is reduced to a matrix $\boldsymbol{D}'$ of size $m$ x $n$ such that
$$M = k_1 m$$
$$N = k_2 n$$
where $k_1, k_2 \in \mathbb{Z}$. Intuitively, choosing a higher $k_1$ and $k_2$ will result in a higher resolution over which the density is evaluated. Selection of $k_1$ and $k_2$ should depend on how large of a region over which motion should be estimated. The upper limits of $k_1$ and $k_2$ are $M$ and $N$, respectively, at which point $\boldsymbol{D}'$ = $\boldsymbol{D}$. Each index $(m, n)$ in $\boldsymbol{D}'$ corresponds to the upper-left corner of a rectangular section of $\boldsymbol{A}$ whose width is $\frac{M}{m}$ and height is $\frac{N}{n}$.
\par Regions of motion are the rectangular sections whose density exceeds a user-defined threshold $\Gamma$, denoting the number of difference pixels over the total number of pixels considered in the rectangular section. For each such region $\boldsymbol{R}$ in $\boldsymbol{A}$ of size $\frac{M}{m}$ x $\frac{N}{n}$,
$$\textnormal{DifferenceDensity}(\boldsymbol{R}) = \frac{\textnormal{number of pixel differences in } \boldsymbol{R}}{\textnormal{number of pixels in } \boldsymbol{R}}$$
\[
\boldsymbol{D}' =
\begin{cases}
1 & \textnormal{DifferenceDensity}(\boldsymbol{R}) > \Gamma\\
0 & \textnormal{otherwise}
\end{cases}
\]
\begin{algorithm}[]
	\smaller
	\label{difference-density-matrix-algorithm}
	\caption{Estimating regions of motion in an image with a difference density matrix: pseudocode for a serial, CPU implementation}

@@

\section*{Acknowledgments}
Placeholder text

>>>>

\documentclass[journal]{IEEEtran}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{algorithm2e}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{listings}
\usepackage{float}
\usepackage{color}
\usepackage{bm}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
	language=C,
	aboveskip=3mm,
	belowskip=3mm,
	showstringspaces=false,
	columns=flexible,
	basicstyle={\small\ttfamily},
	numbers=none,
	numberstyle=\tiny\color{gray},
	keywordstyle=\color{blue},
	commentstyle=\color{dkgreen},
	stringstyle=\color{mauve},
	breaklines=true,
	breakatwhitespace=true,
	tabsize=3
}

@@

\subsection{Separable Convolution}
The filtering necessary in steps (1) and (2) from our algorithm above involve the convolution of a two-dimensional image and a two-dimensional kernel filter of constant width and height. A na\"{\i}ve implementation of two-dimensional convolution would require a double sum across the kernel and image for each individual pixel:
\begin{equation}
	\label{2d-naive-convolution}
	y[m, n] = x[m, n] * h[m, n] = \sum_{i = -\infty}^{\infty} \sum_{j = -\infty}^{\infty} h[i, j] x[m - i, n - j]
\end{equation}
For an image of size $M$ x $N$ and kernel of size $k$, a direct implementation would require $O(MNk^2)$ time, which is infeasible for a real-time implementation. Instead, it is possible to exploit the separable property of certain convolution matrices $h$, e.g. that $h$ can be represented as the product of two one-dimensional matrices $h_1$ and $h_2$. This effectively reduces our two-dimensional convolution of Equation \eqref{2d-naive-convolution} to two separate instances of one-dimensional convolution:
\begin{equation}
	\label{separable-convolution-definition}
	y[m, n] = h_1[n] * (h_2[m] * x[m, n])
\end{equation}
which is implemented as
\begin{equation}
	\label{separable-convolution-formula}
	y[m, n] = \sum_{i = -\infty}^{\infty}h_1[i] \sum_{j = -\infty}^{\infty}h_2[j] x[m - i, n - j]
\end{equation}
For the same $M$ x $N$ input and square kernel of size $k$, a separable implementation reduces the computational complexity to $O(MNk)$. By comparison, an implementation using the FFT would cost $O(MN\log{MN})$. In practice, consider the following results\footnote{Results were obtained by measuring the execution time of a MATLAB script on an Intel Core i7 4790K @ 4.7 GHz.} for a convolution of a $3$ x $3$ kernel with an image of size $2500$ x $2500$:
\begin{table}[H]
	\small
	\centering
	\caption{Comparison of Computational Complexity of Different Convolution Implementation Algorithms}
	\label{convolution-complexity-comparison}
	\begin{tabular}{llll}
	 & Complexity & Execution time (s) \\
	 Direct convolution & $O(MNk^2)$ & $3$ \\
	 Separated convolution & $O(MNk)$ & $3$ \\
	 FFT & $O(MN\log{MN})$ & $3$
	\end{tabular}
\end{table}
As detailed in later sections, we take advantage of the fast computational complexity of separable convolution in our implementation, as all of the filters we apply are separable into one-dimensional matrices.

@@

\subsection{GPU Parallelization and CUDA Implementation}
We begin with a parallelized implementation of separable convolution, since the same convolution operation is applied for both the initial Gaussian low pass filter and the edge detecting Sobel filter. Our parallel implementation involves first independently computing the convolution with the horizontal filter ($h_2$ in Equation \eqref{separable-convolution-definition}), then convolving that result with the vertical filter $h_1$. In computing the output $y[m, n]$ at every index $(i, j)$, we launch a kernel with 16 threads per block, where the total number of blocks along one dimension is equal to the size of that dimension divided by 16. For our 5312 x 2988 test images, this corresponds to 332 blocks in the horizontal direction and 186 blocks in the vertical direction, each with 16 computational threads.
\begin{lstlisting}[
	caption=Grid configuration for the separable convolution device kernel,
	label=grid-configuration
]
#define TX 16
#define TY 16

dim3 block_size(TX, TY);
int bx_horizontal = horizontal_convolution_width/block_size.x;
int by_horizontal = horizontal_convolution_height/block_size.y;
dim3 grid_size_horizontal = dim3(bx_horizontal, by_horizontal);
int bx_vertical = vertical_convolution_width/block_size.x;
int by_vertical = vertical_convolution_height/block_size.y;
dim3 grid_size_vertical = dim3(bx_vertical, by_vertical);

horizontal_convolve<<<grid_size_horizontal, block_size>>>(...);
vertical_convolve<<<grid_size_vertical, block_size>>>(...);
\end{lstlisting}
\par The convolution on the device executes only one serial loop across the dimensions of the filter exactly as it is defined in Equation \eqref{separable-convolution-formula}. This sum is then stored in the output array\footnote{In our implementation, we represent a two-dimensional image as a one-dimensional array by directly concatenating the rows of the matrix into a single array. Thus, for a matrix of size $M$ x $N$, the index $(i, j)$ would correspond to index $Ni + j$ in our one-dimensional array.} according to the block index, block size, and thread index. The implementation is modularized so that any separable filter can be applied to an input matrix.
\begin{lstlisting}[
	caption=Implementation of horizontal and vertical separated convolution,
	label=separable-convolution-implementation
]
__global__ void horizontal_convolve(int *d_out, int *x, int *h, int x_width, int x_height, int h_width, int h_height) {
    const int r = blockIdx.y * blockDim.y + threadIdx.y;
    const int c = blockIdx.x * blockDim.x + threadIdx.x;
    const int i = r * (x_width + h_width - 1) + c;

    int sum = 0;
    for (int j = 0; j < h_width; j++) {
        int p = x_width*r + c - j;
        if (c - j >= 0 && c - j < x_width) {
            sum += h[j] * x[p];
        }
    }
    d_out[i] = sum;
    __syncthreads();
}

__global__ void vertical_convolve(int *d_out, int *x, int *h, int x_width, int x_height, int h_width, int h_height, double constant_scalar) {
	const int r = blockIdx.y * blockDim.y + threadIdx.y;
	const int c = blockIdx.x * blockDim.x + threadIdx.x;
	const int i = r * x_width + c;

    int sum = 0;
    for (int j = 0; j < h_height; j++) {
        int p = x_width*(r - j) + c;
        if (r - j >= 0 && r - j < x_height) {
            sum += h[j] * x[p];
        }
    }
    d_out[i] = (int)(constant_scalar * (double)sum);
    __syncthreads();
}
\end{lstlisting}

@@

\subsection{Motion Area Estimation via a Spatial Difference Density Map}
The difference matrix $\boldsymbol{D}$ determined with Algorithm \ref{difference-matrix-algorithm} is effectively a maximally precise estimate of motion area (e.g., to the resolution of a single pixel, since $\boldsymbol{D}$ is computed on a pixel-by-pixel basis). However, in practice, it is more useful to consider general regions of the image around which motion exists. For this reason, we build a density map of the matrix $\boldsymbol{D}$ using equal-sized rectangular sections of the input image that approximates the density of the difference per unit area (units pixels$^2$). Thus, the original $M$ x $N$ difference matrix $\boldsymbol{D}$ is reduced to a matrix $\boldsymbol{D}'$ of size $m$ x $n$ such that
$$M = k_1 m$$
$$N = k_2 n$$
where $k_1, k_2 \in \mathbb{Z}$. Intuitively, choosing a higher $k_1$ and $k_2$ will result in a higher resolution over which the density is evaluated. Selection of $k_1$ and $k_2$ should depend on how large of a region over which motion should be estimated. The upper limits of $k_1$ and $k_2$ are $M$ and $N$, respectively, at which point $\boldsymbol{D}'$ = $\boldsymbol{D}$. Each index $(m, n)$ in $\boldsymbol{D}'$ corresponds to the upper-left corner of a rectangular section of $\boldsymbol{A}$ whose width is $\frac{M}{m}$ and height is $\frac{N}{n}$.
\par Regions of motion are the rectangular sections whose density exceeds a user-defined threshold $\Gamma$, denoting the number of difference pixels over the total number of pixels considered in the rectangular section. For each such region $\boldsymbol{R}$ in $\boldsymbol{A}$ of size $\frac{M}{m}$ x $\frac{N}{n}$,
$$\textnormal{DifferenceDensity}(\boldsymbol{R}) = \frac{\textnormal{number of pixel differences in } \boldsymbol{R}}{\textnormal{number of pixels in } \boldsymbol{R}}$$
\[
\boldsymbol{D}' =
\begin{cases}
1 & \textnormal{DifferenceDensity}(\boldsymbol{R}) > \Gamma\\
0 & \textnormal{otherwise}
\end{cases}
\]
\begin{figure}[]
	\centering
	\includegraphics[width=0.45\textwidth]{desk_difference_density.jpg}
	\caption{Spatial difference density matrix $\boldsymbol{D}'$ (visualized in three dimensions) of Figure \ref{desk-difference}, as determined by Algorithm \ref{difference-density-matrix-algorithm} with parameters $m = 10$, $n = 6$, and $\Gamma = 0.01$}
    \label{desk-difference-density}
\end{figure}
\begin{figure}[]
	\centering
	\includegraphics[width=0.45\textwidth]{desk_motion_area.jpg}
	\caption{Final estimate of motion area of Figures \ref{desk-1-original} and \ref{desk-2-original} based on the density map of Figure \ref{desk-difference-density}}
    \label{desk-difference-density}
\end{figure}
\begin{algorithm}[]
	\smaller
	\label{difference-density-matrix-algorithm}
	\caption{Estimating regions of motion in an image with a difference density matrix: pseudocode for a serial, CPU implementation}

@@

\section*{Team}
\begin{itemize}
	\item \textbf{Emilio Del Vechhio}, Electrical and Computer Engineering `18
	\item \textbf{Kevin Lin}, Electrical and Computer Engineering `18
	\item \textbf{Senthil Natarjan}, Electrical and Computer Engineering `17
\end{itemize}

\section*{Acknowledgments}
\begin{itemize}
	\item \textbf{CJ Barberan}, Ph.D. student, Rice University Electrical and Computer Engineering---\textit{for mentoring the entire team in the conception, development, and extensive debugging of our project}.
\end{itemize}
